\Large\textbf{Advection versus diffusion in brain ventricular transport}

### 2025-04-02 paper/main.tex
Update abstract intro w/HH

\begin{abstract}
Cerebrospinal fluid (CSF) is integral to brain function, as the fluid
provides mechanical support for the brain and the flow of CSF helps
distribute nutrients, neurotransmitters and metabolites throughout the
central nervous system. CSF flow is partly driven by beating of motile
cilia, hair-like structures that line the ependymal cell layer of the
walls of the brain ventricles. Despite the physiological importance of
CSF flow, the underlying mechanisms of cilia-mediated flow and
transport in the brain ventricles remain yet to be comprehensively
resolved. This study analyzes and evaluates the role of motile cilia
in brain intraventricular CSF flow and transport. The geometry of the
zebrafish larval brain ventricles, for which detailed knowledge of
cilia properties and CSF motion velocities and frequencies exists, was
used to develop finite element methods for modelling flow and
transport. The computational model is validated by in-vivo experiments
that monitor transport of a photoconvertible protein secreted in the
brain ventricles. Our results show that while cilia contribute to
advection of large particles, diffusion plays a significant role in
the transport of small molecules \mer{such as proteins}. We also
demonstrate how cilia location and the geometry of the ventricular
system impact solute distribution. Altogether, this work presents a
computational framework that can be applied to other ventricular
systems, together with new concepts of how molecules are transported
within the brain and its ventricles.

@@

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure gradients generated by the
cardiac cycle, respiration, neural activity, and \mer{bodily}
movements~\cite{Brodal2016, del2010ependymal, Ringers2020Role, Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
macaulay2022cerebrospinal, mestre2018flow}.
Cilia are hair-like structures found in a vast diversity of
biological organisms and in several human organs,
including the brain~\cite{DGama2025MotileBrain,
Fulford1986Muco-ciliaryLung,
Pacherres2022CiliaryProduction, Jahn1972LocomotionProtozoa,
Brennen1977FluidFlagella, Ringers2020Role, spassky2013motile,
Pellicciotta2020Cilia, mitchell2007evolution,
Tsukita2012CoordinatedFeet, BLAKE1974MechanicsMotion,
Reiten2017Motile-Cilia-MediatedComputations, Thouvenin2020OriginCanal}. 
The ventricular surface of the brain, known as the ependyma, is lined by
motile cilia that beat in a coordinated pattern at frequencies of
10--40 Hz to mediate fluid flow~\cite{mitchell2007evolution,
Ringers2020Role, spassky2013motile, Ringers2023NovelEpithelia, roth2025structure}. 
Given their microscopic size, ranging from
5 to 15 microns, cilia contribute primarily to flow near the
ependyma in the human brain~\cite{Siyahhan2014FlowVentricles, Olstad2019CiliaryDevelopment,
Ringers2020Role, Faubel2016Cilia-basedVentricles}. In contrast, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled, mestre2018flow, daversin2020mechanisms}.

% MER: Note, we do not write "endeavors" nor "delve". (GenAI does.)

% MER: Note "Failed to identify" indicates that someone tried but
% failed. It is a very effective way of pissing said authors (often
% reviewers or potential future collaborators) off if they just didn't
% try/that was not their target.

% I3: What do we not know about the problem and what do we want to
% know? What are the precise research questions we want to address?
% Refer to specific studies and what specifically we do not know.
To date, it remains poorly understood how motile cilia contribute to 
the overall fluid flow and solute transport within the brain.
Experimental approaches face technical challenges: the microscopic size and relatively 
high beating frequencies of the cilia~\cite{ocallaghan2012analysis, mahuzier2018ependymal,
Jeong2022MeasurementTelencephalon, Olstad2019CiliaryDevelopment, spassky2013motile, DGama2021Diversity},
as compared to other fluid dynamic processes in the brain that are slower and span larger
spatial scales~\cite{Vinje2019RespiratoryMeasurements, eide2024functional, fultz2019coupled, mestre2018flow},
make cilia difficult to monitor in vivo in humans.
Until now, most studies have analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles~\cite{Faubel2016Cilia-basedVentricles, Sawamoto2006NewBrain, ibanez2003beat};
however, they \mer{did not study} the role of other contributions to CSF
flow and the accompanying transport, such as pressure
gradients associated with respiration and cardiac pulsations. Moreover, while MRI technology
can measure CSF motion in humans~\cite{feinberg1987human, battal2011cerebrospinal,
yamada2015current}, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain~\cite{Siyahhan2014FlowVentricles}. 
These technical limitations of studying cilia-mediated flow and transport make 
mathematical modeling and computational investigations an attractive complement to experimental
studies. Previous simulation studies have focused on modeling the interaction
between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}, but also 
other applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays, roth2025structure},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella}, and
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.
However, we still lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport within the brain.
This is of particular importance since disorders such as hydrocephalus are observed
in animal models and humans with deficient motile cilia~\cite{Afzelius2004CiliaRelatedDiseases,
Olstad2019CiliaryDevelopment, Mitchison2017Motile, Ibanez-Tallon2002LossHydrocephalus},
indicating that the cilia play a crucial role for healthy brain function.

>>>>

\begin{abstract}
Cerebrospinal fluid (CSF) is integral to brain function, as the fluid
provides mechanical support for the brain and the flow of CSF helps
distribute nutrients, neurotransmitters and metabolites throughout the
central nervous system. CSF flow is partly driven by beating of motile
cilia, hair-like structures that line the ependymal cell layer of the
walls of the brain ventricles. Despite the physiological importance of
CSF flow, the underlying mechanisms of cilia-mediated flow and
transport in the brain ventricles remain yet to be comprehensively
resolved. This study analyzes and evaluates the role of motile cilia
in brain intraventricular CSF flow and transport. The geometry of the
zebrafish larval brain ventricles, for which detailed knowledge of
cilia properties and CSF motion velocities and frequencies exists, was
used to develop finite element methods for modelling flow and
transport. The computational model is validated by in-vivo experiments
that monitor transport of a photoconvertible protein secreted in the
brain ventricles. Our results show that while cilia contribute to
advection of large particles, diffusion plays a significant role in
the transport of small solutes. We also demonstrate how cilia location
and the geometry of the ventricular system impact solute
distribution. Altogether, this work presents a computational framework
that can be applied to other ventricular systems, together with new
concepts of how molecules are transported within the brain and its
ventricles.

@@

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure
gradients generated by the cardiac cycle, respiration, neural
activity, and body movements~\cite{Brodal2016, del2010ependymal,
  Ringers2020Role, Vinje2019RespiratoryMeasurements,
  Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
  macaulay2022cerebrospinal, mestre2018flow}.  Cilia are hair-like
structures found in a vast diversity of biological organisms and in
several human organs, including the brain~\cite{DGama2025MotileBrain,
  Fulford1986Muco-ciliaryLung, Pacherres2022CiliaryProduction,
  Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella,
  Ringers2020Role, spassky2013motile, Pellicciotta2020Cilia,
  mitchell2007evolution, Tsukita2012CoordinatedFeet,
  BLAKE1974MechanicsMotion,
  Reiten2017Motile-Cilia-MediatedComputations,
  Thouvenin2020OriginCanal}.  The ventricular surface of the brain,
known as the ependyma, is lined by motile cilia that beat in a
coordinated pattern at frequencies of 10--40 Hz to mediate fluid
flow~\cite{mitchell2007evolution, Ringers2020Role, spassky2013motile,
  Ringers2023NovelEpithelia, roth2025structure}.  Given their
microscopic size, ranging from 5 to 15 microns, cilia contribute
primarily to flow near the ependyma in the human
brain~\cite{Siyahhan2014FlowVentricles, Olstad2019CiliaryDevelopment,
  Ringers2020Role, Faubel2016Cilia-basedVentricles}. In contrast,
cardiac, respiratory and delta waves at lower frequencies induce
pulsatile flow at the larger scales in the brain ventricular system,
subarachnoid space, and subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
  Causemann2025, fultz2019coupled, mestre2018flow,
  daversin2020mechanisms}.

% I3: What do we not know about the problem and what do we want to
% know? What are the precise research questions we want to address?
% Refer to specific studies and what specifically we do not know.
To date, it remains poorly understood how motile cilia contribute to 
the overall fluid flow and solute transport within the brain.
Experimental approaches face technical challenges: the microscopic size and relatively 
high beating frequencies of the cilia~\cite{ocallaghan2012analysis, mahuzier2018ependymal,
Jeong2022MeasurementTelencephalon, Olstad2019CiliaryDevelopment, spassky2013motile, DGama2021Diversity},
as compared to other fluid dynamic processes in the brain that are slower and span larger
spatial scales~\cite{Vinje2019RespiratoryMeasurements, eide2024functional, fultz2019coupled, mestre2018flow},
make cilia difficult to monitor in vivo in humans.
Until now, most studies have analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles~\cite{Faubel2016Cilia-basedVentricles, Sawamoto2006NewBrain, ibanez2003beat};
however, they did not study the role of other contributions to CSF
flow and the accompanying transport, such as pressure
gradients associated with respiration and cardiac pulsations. Moreover, while MRI technology
can measure CSF motion in humans~\cite{feinberg1987human, battal2011cerebrospinal,
yamada2015current}, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain~\cite{Siyahhan2014FlowVentricles}. 
These technical limitations of studying cilia-mediated flow and transport make 
mathematical modeling and computational investigations an attractive complement to experimental
studies. Previous simulation studies have focused on modeling the interaction
between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}, but also 
other applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays, roth2025structure},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella}, and
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.
However, we still lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport within the brain.
This is of particular importance since disorders such as hydrocephalus are observed
in animal models and humans with deficient motile cilia~\cite{Afzelius2004CiliaRelatedDiseases,
Olstad2019CiliaryDevelopment, Mitchison2017Motile, Ibanez-Tallon2002LossHydrocephalus},
indicating that the cilia play a crucial role for healthy brain function.
